---
title: "Data Exploration"
author: "Greta Dertwinkel"
date: "05-03-25"
abstract: "In this document, we analyze the Titanic dataset to understand the fate of its passengers."

output:
  html_document:
    code_folding: show
    toc: true
---

# Introduction

Lindner & Erixson (2012, PNAS) provide data on 18 maritime disasters since 1855, including the Titanic. Their dataset includes over 15,000 cases and 17 variables (available for free at PNAS: <https://www.pnas.org/content/109/33/13220>). The Titanic is one of two disasters (the HMS Birkenhead is the other) in which female survival exceeded male survival. From the abstract: “Women have a distinct survival disadvantage compared with men. Captains and crew survive at a significantly higher rate than passengers. Taken together, our findings show that human behavior in life-and-death situations is best captured by the expression ‘every man for himself.’”

Is that the case? Let's try to understand more about the fate of the passengers.

## Methods:

In the following document, we will explore the Titanic dataset. This dataset provides information on the passengers of the ocean liner ['Titanic'](https://en.wikipedia.org/wiki/Sinking_of_the_Titanic). We adapt content from [Regression Modelling Strategies by Frank Harrell](https://hbiostat.org/rmsc/intro). Along the way we will learn to use a few useful functions from the RMS and Hmisc packages, while applying some best practices in project management and reporting.

## Descriptive Statistics

We will begin with some descriptive statistics:

```{r load-libraries, include=FALSE}
# List of required packages
packages <- c("ggplot2", "Hmisc", "rms", "tidyverse", "kableExtra", "units")

# Install missing packages
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
  }
}

# Apply function to all packages
invisible(lapply(packages, install_if_missing))

# load libraries
library("ggplot2")
library("Hmisc")
library("rms") 
library("tidyverse")
library("kableExtra")
library("units")
```

```{r setup, echo=FALSE}
# Set global chunk options for consistency
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Print output in html format
options(prType='html')  

# Load in data and keep only the variables of interest
titanic <- readRDS("titanic3.rds") %>%
  select('pclass', 'survived', 'age', 'sex', 'sibsp', 'parch') %>%
  mutate(age = as.numeric(age)) %>%  # Ensure 'age' is numeric
  mutate(age = set_units(age, "years")) %>% #display age in years
  mutate(age = as.numeric(age)) # Ensure 'age' is numeric
```

```{r data-summary}
# Display a preview of the dataset
kable(head(titanic), caption = "Preview of Titanic Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Check data strcuture
str(titanic)

# Compute summary statistics
summary_stats <- summary(titanic)

# Display summary statistics in a table
kable(summary_stats, caption = "Summary Statistics of Titanic Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Survival Rates

Here we look at univariable summaries of survival rates.

```{r survival-plot}
# Create datadist object for use with 'rms' functions
dd <- datadist(titanic)
options(datadist = 'dd')

# Compute summary statistics using the 'summary' function from 'rms'
s <- summary(survived ~ age + sex + pclass +
               cut2(sibsp, g = 4) + cut2(parch, g = 4), 
             data = titanic)

# Convert 'summary' output into a dataframe for better visualization
s_df <- as.data.frame(s)

# Increase margins before plotting
par(mar = c(5, 10, 4, 2))  # Adjust bottom, left, top, right margins

# Plot the survival rates
plot(s, 
     main = "Variable Distributions and Impact on Survival",
     subtitles = FALSE, 
     cex.main = 1.2,  # Adjust title size
     cex.lab = 0.9,   # Reduce label size
     cex.axis = 0.8,  # Reduce axis text size
     las = 1)         # Rotate text for better readability
```

## Exploring Trends

In this section, we will explore the dataset and try to identify trends in survival rates.

We use nonparametric regression (loess) estimates of the relationship between age and survival probability.

```{r survival-age}
# Create a plot of survival rate across age
yl <- ylab(NULL) # create object yl
plot_age <- ggplot(titanic, aes(x=age, y=survived)) +
  histSpikeg(survived ~ age, lowess=TRUE, data=titanic, histcol = "purple") +
  ylim(0,1) + yl +
  labs( x = "Age of Passengers", 
        y = "Survival Probability",
        title = "Survival Rate of Passengers across Age",
        subtitle = "Survival Plot with LOESS Smoothed Line",
        caption = "Data Source: Titanic Dataset") + # set labels for the plot
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        panel.grid.minor = element_line(color = "gray", linetype = "dotted")) #theme adjustments
  

# Plot
print(plot_age)
```

What we observe is that survival rates decrease as a funciton of age.

Let's look at whether this relationship differs by gender:

```{r survival-gender}
# Create a plot of survival rate across age
plot_gender <- ggplot(titanic, aes(x=age, y=survived, color=sex)) +
  histSpikeg(survived ~ age + sex, lowess=TRUE, data=titanic, histcol = "purple") +
  ylim(0,1) + yl +
  labs( x = "Age of Passengers", 
        y = "Survival Probability",
        title = "Survival Rate of Passengers across Age and Gender",
        subtitle = "Survival Plot with LOESS Smoothed Line",
        caption = "Data Source: Titanic Dataset",
        size ="Gender") + # set labels for the plot 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        panel.grid.minor = element_line(color = "gray", linetype = "dotted")) # theme adjustments
  

# Plot
print(plot_gender)
```

And class:

```{r survival-class}
# Set the scale for the plot lines
b  <- scale_size_discrete(range=c(.1, .85))

# Create a plot of survival rate across age
plot_class <- ggplot(titanic, aes(x=age, y=survived, size=pclass)) +
  histSpikeg(survived ~ age + pclass, lowess=TRUE, data=titanic) + b +
  ylim(0,1) + yl +
  labs( x = "Age of Passengers", 
        y = "Survival Probability",
        title = "Survival Rate of Passengers across Age and Class",
        subtitle = "Survival Plot with LOESS Smoothed Line",
        caption = "Data Source: Titanic Dataset",
        size = "Class") + # set labels for the plot 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        panel.grid.minor = element_line(color = "gray", linetype = "dotted")) # theme adjustments

# Plot
print(plot_class)
```

Let's combine all this information:

```{r combine-plots}
# Create a plot of survival rate across age
plot_comb <- ggplot(titanic, aes(x=age, y=survived, color = sex, size=pclass)) +
  histSpikeg(survived ~ age + sex + pclass, lowess=TRUE, data=titanic) + b +
  ylim(0,1) + yl +
  labs( x = "Age of Passengers", 
        y = "Survival Probability",
        title = "Survival Rate of Passengers across Age, Class and Sexuality",
        subtitle = "Survival Plot with LOESS Smoothed Line",
        caption = "Data Source: Titanic Dataset",
        size = "Class") + # set labels for the plot 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "italic"),
        plot.caption = element_text(size = 10, face = "italic"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        panel.grid.minor = element_line(color = "gray", linetype = "dotted")) + # theme adjustments
  guides(color = guide_legend(title = "Gender"))

# Plot
print(plot_comb)

# Combine all four plots
gridExtra::grid.arrange(plot_age, plot_gender, plot_class, plot_comb, ncol=2)   # combine 4
```

Run a quick comparison:

```{r quick-comparison}
# Plot 1: Smooth curve of survival vs age (no grouping)
p1 <- ggplot(titanic, aes(x = age, y = survived)) +
      geom_smooth(method = "loess", color = "darkblue", size = 1) +  # Adjusted color and line thickness
      ylim(0, 1) +
      labs(title = "Survival vs Age (Overall)", y = "Survival Probability", x = "Age of Passengers") +
      theme_minimal() +
      theme(plot.title = element_text(size = 14, face = "bold", color = "#2C3E50"),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))

# Plot 2: Survival vs Age by Sex
p2 <- ggplot(titanic, aes(x = age, y = survived, color = sex)) +
      geom_smooth(method = "loess", size = 1) +
      ylim(0, 1) +
      labs(title = "Survival vs Age by Sex", y = "Survival Probability", x = "Age of Passengers", color = "Sex") +
      theme_minimal() +
      theme(plot.title = element_text(size = 14, face = "bold", color = "#2C3E50"),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))

# Plot 3: Survival vs Age by Passenger Class 
p3 <- ggplot(titanic, aes(x = age, y = survived, size = as.factor(pclass))) +
      geom_smooth(method = "loess", color = "darkgreen", size = 1) +
      ylim(0, 1) +
      labs(title = "Survival vs Age by Class", y = "Survival Probability", x = "Age of Passengers", size = "Passenger Class") +
      theme_minimal() +
      theme(plot.title = element_text(size = 14, face = "bold", color = "#2C3E50"),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))

# Plot 4: Survival vs Age by Sex and Passenger Class
p4 <- ggplot(titanic, aes(x = age, y = survived, color = sex, size = as.factor(pclass))) +
      geom_smooth(method = "loess") + b +
      ylim(0, 1) +
      labs(title = "Survival vs Age by Sex and Class", y = "Survival Probability", x = "Age of Passengers", color = "Sex", size = "Passenger Class") +
      theme_minimal() +
      theme(plot.title = element_text(size = 14, face = "bold", color = "#2C3E50"),
            axis.title = element_text(size = 12),
            axis.text = element_text(size = 10))

# Arrange the plots in a grid
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```

The top-left panel shows unstratified estimates, while the other panels stratify by different variables.

## Discussion

Our findings confirm that passenger class, gender, and age had significant impacts on survival. Women had the highest survival rates, but the picture bicomes more nuanced when we stratify by other variables of interest.

## Bibliography
Elinder, M., & Erixson, O. (2012). Gender, social norms, and survival in maritime disasters. *Proceedings of the National Academy of Sciences*, *109*(33), 13220–13224. <https://doi.org/10.1073/pnas.1207156109>
